---
title: "NCAA Basketball Analysis"
author: "Saul, Jackson, Cody"
date: "4/23/24"
output:
  ioslides_presentation: default
  beamer_presentation:
    theme: Madrid
  slidy_presentation: default
fontsize: 9pt
graphics: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = FALSE, warning = FALSE, fig.align = "center", fig.height = 4, fig.width = 6,  out.width = "75%")
library(knitr)
library(tidyverse)
library(moderndive)
library(GGally)
library(tidyverse)
library(caret)
library(rpart)  #library for CART
library(rpart.plot)

cbb <- read_csv("cbb.csv")
cbb=cbb%>%
  mutate(win_perc=W/G)
```

## Outline

In this lecture, we will...

\pause

-   Introduce our what our data contains and interesting findings

-   Discuss Multiple Linear Regression

-   Discuss Regression Tree

-   Discuss Polynomial Regression

-   Compare findings

# Our Data

## Columns

-   This data contained 24 different columns of data from team name,
    games played and won, two point and three point shooting percentages
    for themselves and for their match up. There was only four
    categorical variables: Team name, Conference, Seed and Postseason
    result which was where they got knocked out.

\pause

-   Our most important variables were BARTHAG (Power Rating), Win
    percentage, Power 5, Number of games played and Number of games won.

    \pause

    -   Win Percentage (win_perc) and Power 5 (power_5) were created
        variables by us in order to achieve our goal

## Our Response Variable...

\pause

-   Our response variable was Win Percentage as the number of games
    played differed from team to team

    -   Win Percentage was found by simply dividing the number of games
        won by the total number of games played

        \pause

-   Our main goal is to determine how the power rating of a team effects
    its win percentage across various seasons. Then to determining what
    model is best for predicting this percentage.

    \pause

-   Our data was also divided by Power 5 which determined whether a team
    was in a power five conference which are the biggest and most
    competitive conferences in the nation

# Linear Regression

## Finding relative variables

```{r echo=FALSE, fig.height=3}

pairs=cbb%>%
  select("3P_O","2P_O","3P_D","2P_D","BARTHAG","win_perc")
ggpairs(pairs)
```

-   \pause

    -   Our variable with the most correlation to our response variable
        was BARTHAG

        -   BARTHAG is a teams power rating, in other words, the chance
            of beating an average D1 team

    ## Simple Linear Regression

    ```{r echo=FALSE, fig.height=3}
    #Training and Testing Data
    trainingInd=sample(1:3523,2466)
    train_data=cbb[trainingInd,]
    test_data=cbb[-trainingInd,]

    #Fitting linear model
    lin_mod=lm(win_perc~BARTHAG,data=train_data)
    train_data%>%
      ggplot(aes(x=BARTHAG,y=win_perc))+
      geom_point()+
      geom_smooth(method="lm")+
      theme_minimal()+
      labs(x="Power Rating",y="Win Percentage",title="Win % vs BARTHAG (Power Rating)")
    ```

    \pause

    The relationship is decently significant and compared to the rest of
    our variable it does correlate the most as the more a teams power
    rating increases, their win percentage does tend to increase as well
    but not by a significant amount

## Multiple Linear Regression

```{r include=FALSE}
"Creates Power 5 converence variable"
#%in% checks if the conference name matches ANY of the values in the list
#Returns TRUE if the team is in ACC, B10, B12, P12, or SEC
#Returns FALSE for all other conferences
train_data$power_5 = ifelse(train_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")

#If TRUE (team is in Power 5 conference) then assigns "Power 5"
#If FALSE (team is not in Power 5 conference) then assigns "Non-Power 5"
# Create power_5 variable in test data
test_data$power_5 = ifelse(test_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")
parallel_mod = lm(win_perc ~ BARTHAG + power_5, data=train_data)

trctrl_mlr <- trainControl(method = "repeatedcv", number = 5, repeats = 10)

mlr<-  train(win_perc ~ BARTHAG + power_5, 
                   data = train_data, 
                   method = "lm",
                   trControl=trctrl_mlr)
```

```{r echo=FALSE,fig.height=3}

train_data %>%
  ggplot(aes(x=BARTHAG, y=win_perc)) +
  geom_point(alpha=.2,aes(color=power_5)) +
  geom_parallel_slopes(aes(color=power_5),linewidth=1.5) +
  theme_bw() +
  labs(x="Power Rating ", y="Win Percentage ", title="Parallel Lines Model: Win % vs BARTHAG by Conference Type",color="Conference Type")


```

Teams in a power five conference had a higher average power rating
compared to teams not in a power five conference. However, the more
teams in a power conference had a lower win percentage, most likely due
to being in a more competitive conference overall.

The RMSE for this model is 0.1088913

```         
```

# Regression Tree

## The Best Tree

\pause

-   RMSE measures how well the regression tree predicts win percentage.

-   As cp increases, RMSE also increases, indicating that pruning the
    tree too much reduces predictive accuracy.

-   This helps us tune our regression tree by choosing the best cp value
    that minimizes RMSE

```{r,fig.height=3}
library(caret)
library(rpart)  #library for CART
library(rpart.plot)

train_data <- na.omit(train_data)

trctrl_1 <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

caretTree  <-  train(win_perc ~ BARTHAG + power_5 + ADJ_T + SEED + WAB, 
                     data = train_data, 
                     method = "rpart",
                     trControl=trctrl_1,
                     tuneGrid = expand.grid(cp=seq(0.001, 0.1, 0.001))
)
plot(caretTree)


```

## The Best Tree

\small

-   Each split represents a decision rule that helps predict win
    percentage based on different variables.

-   The deeper the tree, the more splits, which means more predictions.

-   The tree confirms that WAB and BARTHAG are crucial for predicting
    win percentage, aligning with our expectations.

```{r,fig.height=3}
library(caret)
library(rpart)  #library for CART
library(rpart.plot)

train_data <- na.omit(train_data)

trctrl_1 <- trainControl(method = "repeatedcv", number = 10, repeats = 10)

caretTree  <-  train(win_perc ~ BARTHAG + power_5 + ADJ_T + SEED + WAB, 
                     data = train_data, 
                     method = "rpart",
                     trControl=trctrl_1,
                     tuneGrid = expand.grid(cp=seq(0.001, 0.1, 0.001))
)

rpart.plot(caretTree$finalModel)

```

## LM

\pause

-   Some variables, like WAB, BARTHAG, and Power 5, were highly
    significant, while others, like ADJ_T and some SEED values, were
    not.

```{r}
library(caret)
library(rpart)  #library for CART
library(rpart.plot)

trctrl_2 <- trainControl(method = "repeatedcv", number = 5, repeats = 10)

caretLM<-  train(win_perc ~ BARTHAG + power_5 + ADJ_T + SEED + WAB, 
                 data = train_data, 
                 method = "lm",
                 trControl=trctrl_2
)

summary(caretLM)
```

## Comparing The Models

\pause

-   Intercept Model Performs Better

```{r}
trctrl_2 <- trainControl(method = "repeatedcv", number = 5, repeats = 10)

caretLM<-  train(win_perc ~ BARTHAG + power_5 + ADJ_T + SEED + WAB, 
                 data = train_data, 
                 method = "lm",
                 trControl=trctrl_2
)

caretLM$results


```

## Comparing The Models

\small

-   CP Model is Worse in Every Metric

```{r}
trctrl_2 <- trainControl(method = "repeatedcv", number = 5, repeats = 10)

caretLM<-  train(win_perc ~ BARTHAG + power_5 + ADJ_T + SEED + WAB, 
                 data = train_data, 
                 method = "lm",
                 trControl=trctrl_2
)



caretTree$results[caretTree$results$cp==caretTree$bestTune[1,1], ]
```

# Polynomial Regression

## Which degree should we use?

\small

-   To figure out which degree polynomial to use, we calculated RMSE for
    each degree

-   3rd degree seems to be the best suitable, after that the RMSE
    doesn't change much

```{r echo = F, eval = T}
library(GGally)
library(tidyverse)
library(caret)
cbb <- read_csv("cbb.csv")
cbb=cbb%>%
  mutate(win_perc=W/G)
pairs=cbb%>%
  select("3P_O","2P_O","TORD","ORB","3P_D","2P_D","BARTHAG","win_perc")
trainingInd=sample(1:3523,2466)
train_data=cbb[trainingInd,]
test_data=cbb[-trainingInd,]
cbb <- cbb %>%
  mutate(power_5 = case_when(
    CONF %in% c("ACC", "B10", "B12", "P12", "SEC") ~ "Power 5",
    TRUE ~ "Non-Power 5"
  ))
train_data$power_5 = ifelse(train_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")

#If TRUE (team is in Power 5 conference) then assigns "Power 5"
#If FALSE (team is not in Power 5 conference) then assigns "Non-Power 5"
# Create power_5 variable in test data
test_data$power_5 = ifelse(test_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")


train_data_complete <- train_data %>%
  drop_na()  # Remove any rows with NA values

# also clean the test data
test_data_complete <- test_data %>%
  drop_na()

# Create models with different polynomial degrees
poly_models <- list()
rmse_values <- numeric(10)

for (degree in 1:10) {
  # Create model predicting win_perc from BARTHAG
  poly_models[[degree]] <- lm(win_perc ~ poly(BARTHAG, degree, raw = TRUE) +
                                power_5 + ADJ_T + SEED + POSTSEASON,
                              data = train_data)

  # Calculate RMSE
  rmse_values[degree] <- sqrt(mean(poly_models[[degree]]$residuals^2))
}

# Compare models
comparison <- data.frame(
  Degree = 1:10,
  RMSE = rmse_values
)


# Plot RMSE by polynomial degree
plot(comparison$Degree, comparison$RMSE, type = "b",
     xlab = "Polynomial Degree", ylab = "RMSE",
     main = "RMSE by Polynomial Degree")
```

## Visualizing the Polynomial Regression

-   Here is a graph of the 3rd degree polynomial

-   Cross Validated RMSE: 0.08895640

```{r echo = F, eval = T, fig.height=4, out.width = "65%"}
library(GGally)
library(tidyverse)
library(caret)
cbb <- read_csv("cbb.csv")
cbb=cbb%>%
  mutate(win_perc=W/G)
pairs=cbb%>%
  select("3P_O","2P_O","TORD","ORB","3P_D","2P_D","BARTHAG","win_perc")
trainingInd=sample(1:3523,2466)
train_data=cbb[trainingInd,]
test_data=cbb[-trainingInd,]
cbb <- cbb %>%
  mutate(power_5 = case_when(
    CONF %in% c("ACC", "B10", "B12", "P12", "SEC") ~ "Power 5",
    TRUE ~ "Non-Power 5"
  ))
train_data$power_5 = ifelse(train_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")

#If TRUE (team is in Power 5 conference) then assigns "Power 5"
#If FALSE (team is not in Power 5 conference) then assigns "Non-Power 5"
# Create power_5 variable in test data
test_data$power_5 = ifelse(test_data$CONF %in% c("ACC", "B10", "B12", "P12", "SEC"), "Power 5", "Non-Power 5")


train_data_complete <- train_data %>%
  drop_na()  # Remove any rows with NA values

# also clean the test data
test_data_complete <- test_data %>%
  drop_na()

ggplot(train_data, aes(x = BARTHAG, y = win_perc, color = power_5)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = TRUE) +
  theme_minimal() +
  labs(
    title = "Polynomial Regression: Win % vs BARTHAG²",
    x = "BARTHAG",
    y = "Win Percentage",
    color = "Conference Type"
  )
```

\normalsize

-   Power 5 teams on average have a much higher power rating than
    Non-Power 5 teams, even with win percentage being relatively close
    between the two

# Conclusion

## Results

```{r}

```
